---
title: "<Protocol name> NAb Processing"
shorttitle: "<Protocol name> NAb"
from:
 - name: Your name
   email: <your_id>@fredhutch.org
includesummary: yes
summary: 
  NAb processing for the <Protocol name> study 
---

<!--
Program Name: preprocess_nab.Rmd
Creation Date: 1/1/2010
Full name of author: Your name
Project or Protocol: <Protocol name>
Purpose or description of program: NAb data processing
Location of program: git@github.com/FredHutch/<protocol name>.git
Location of input data: git@github.com/FredHutch/<protocol name>.git
-->

This report documents the <Protocol name> NAb data processing.

# Preliminaries
We first source our external functions and global variables, load libraries, and set options.

```{r 'prelim', echo=TRUE, eval=TRUE}

# Libraries
suppressPackageStartupMessages({
  library(knitr)
  library(data.table)
  library(xtable)
  library(readxl)
  library(ggplot2)
  library(kableExtra)}
)

# Functions
sys.source("functions.R"), envir=topenv())
sys.source("qc_functions.R"), envir=topenv())
sys.source("protocol_specific.R"), envir=topenv())
```

# Processing
Next we read in the qdata. The only time point in the qdata file is visitno 30, which appears to be week 30.

```{r 'read_data'}
nab <- rbindlist(lapply(nab_qfilenames,
                        function(x){fread(file.path("..", "inst", "extdata", x))[, sourcefile := x]}),
                 use.names=TRUE, fill=TRUE)
setnames(nab, names(nab), tolower(names(nab)))

# add Week
nab[, Week := visitno]
```

```{r 'rx', eval=runrx, echo=runrx}
# merge group info
cat("# Treatment Assignment\nTreatment assignment information is merged in.\n")
# merge in treatment info
setkey(nab, ptid)
nab <- merge(nab, rx, all.x=TRUE)
kableit(setkey(nab[, .N, .(group)], group), caption = "Record count by group")
kableit(setkey(nab[, .N, .(ptid, group)], group, ptid), caption = "Record count by group and ptid")
```

```{r 'tier'}
# merge tier/clade
cat("# Tier/Clade\nTier and clade information is merged in.\n")
setkey(nab, isolate)
nab <- merge(nab, tier, all.x=TRUE) # tier is created in protocol_specific.R
kableit(setkey(nab[, .N, .(isolate, tier, clade)], isolate), caption = "Record count by isolate")
```


# Response call
```{r 'response_call'}
nab[method=="Dilution", titer_mod := nabTiter(titer, "dilution")]
nab[method=="Concentration", titer_mod := nabTiter(titer, "Concentration")]
nab[, response := nabResponse(titer)]

formula <- as.formula(paste(paste(setdiff(names(nab), c("titer", "titer_mod", "response", "poscrit")), collapse = " + "), "~ poscrit"))
nab <- dcast(nab, formula, value.var = c("titer", "titer_mod", "response"))
setkey(nab, celltype, isolate, ptid, visitno)
```

Some summaries of the unaveraged data follow.

## Analytes and Cell Types
```{r}
kableit(nab[, .N, .(isolate, celltype)])
```

## Time points represented in the data, with record count
```{r}
kableit(nab[, .N, Week])
```

## Range of Titer Values
```{r}
kableit(data.table(Statistic = names(summary(nab$titer_mod_50)), Value=as.numeric(summary(nab$titer_mod_50))), caption = "Summary of 50% positive criteria titers")
kableit(data.table(Statistic = names(summary(nab$titer_mod_80)), Value=as.numeric(summary(nab$titer_mod_80))), caption = "Summary of 80% positive criteria titers")
```

## Relationship of titer_mod_50 to titer_mod_80
```{r}
ggplot(nab, aes(x = titer_mod_50, y = titer_mod_80)) +
  geom_point(na.rm=TRUE, show.legend=TRUE)
```

# CDS variables
```{r}
nab[, cds_prot              := "cvd465"]
nab[, cds_sub_prot          := NA]
nab[, cds_subject_id        := paste(cds_prot, ptid, sep = ' ')]
nab[, cds_specimen_type     := 'serum']
nab[, cds_vaccine_matched   := 0]
nab[, cds_response_id50_std := response_50]
nab[, cds_response_id80_std := response_80]
nab[, cds_slope             := log(4)/(log(titer_mod_80)-log(titer_mod_50))]
nab[, cds_lab_code          := ifelse(labid == "MS", "Seaman",
                                      ifelse(labid %in% c("DM","DU"), "Montefiori", labid))]
```

# Output
```{r 'output'}
# separate serum into direct binding dilution and titration dilutions
setkey(nab, ptid, Week)
saveObj("nab", "nab")
```

## Reproducibility Tables

```{r echo=FALSE, eval=TRUE}
path <- "reproducibility.R")
read_chunk(path)
```

```{r repro, echo=FALSE, eval=TRUE}
<<reprotab>>
```

```{r copySourcedFiles, eval = FALSE, echo = FALSE}
# this copies any files that are sourced in this Rmd to inst/doc/ and vignettes/ so that the package vignettes build properly at installation
# this is new code, and may still be a little buggy or not function as expectged, particularly on Windows
thisfile <- knitr::current_input()
con <- file(thisfile, open = "r")
file <- readLines(con)
m1 <- unname(sapply(file, function(x){grepl('sys.source(.*)', x) & !grepl('unname', x)}))
m2 <- unname(sapply(file, function(x){grepl('read_chunk(.*)', x) & !grepl('unname', x)}))
filesToCopy1 <- sapply(strsplit(file[m1], split = '"'), function(x) x[2])
filesToCopy2 <- sapply(strsplit(file[m2], split = '"'), function(x) x[2])
filesToCopy <- c(filesToCopy1, filesToCopy2)
for(f in filesToCopy){
  file_copy(f, paste0(c("../inst/doc/", "../vignettes/"), f), overwrite = TRUE)
}
```

